---
title: "Negative binomial mixture branching process model of transmission"
header-includes:
   - \usepackage{amsmath}
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(BellB)
#library(actuar) #simulate mixture branching process
```

```{r global-options, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r}
R0fn<-function(p, R0D, R0A){
R0<- p*R0D+(1-p)*R0A
}

R0adel<-function(R0D, delta){
  R0A<-R0D + delta
}


NegBinomGen<-function(s, R0, k){
  #returns s = Q(s), where Q(s) = pgf of negative binomial
  gen<-1/(1 + (R0/k)*(1 - s))^k -s #solve prob generating fn = s for s. Note R0 > 1 otherwise prob of extinction = 1
}

NegBinomMixtureGen<-function(s, p, R0D,k, delta){
  #R0A = R0D + delta
  R0A<-R0D + delta
  gen<-p/(1 + (R0A/k)*(1 - s))^k+(1-p)/(1 + (R0D/k)*(1 - s))^k -s
}


mean.chain<-function(R0){ 
  #Argument R0 can be R0 if R0<1, if R0 >1, compute R0*
  meanchain<-1/(1-R0)
}

#variance of offspring function
varNegBinomoffspring<-function(R0,k){ 
  v<-R0*(1+R0/k)
}

#variance of offspring function
varNegBinommixtureoffspring<-function(p, R0D,k, delta){ 
  #variance is concave up quadratic function of p
  v<-R0D*(1+R0D/k) + p*delta*(1 + delta*(1-p) + (2*R0D+delta)/k)
}

NegBinomR0star<-function(R0=1.5, k){
  #R0 needs to be bigger than 1
  #R0* = Q'(z) evaluated at prob of extinction z < 1
  extprob<-uniroot(NegBinomGen, c(0, 0.99), R0=R0, k=k)$root
  R0star<-R0*(1+(R0/k)*(1-extprob))^(-k-1)
}

NegBinomMixR0star<-function(p, R0D,k, delta){
  R0A<-R0D + delta
  extprob<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p, R0D=R0D, k=k, delta=delta)$root
  R0star<-p*R0A*(1+(R0A/k)*(1-extprob))^(-k-1)+(1-p)*R0D*(1+(R0D/k)*(1-extprob))^(-k-1)
}


variance.chainNegBinom<-function(R0,k){ 
  #Argument: R0 > 1, dispersion parameter k
  #Find variance of chain conditioned on extinction
  #do i need a different function for R0<1??
  extprob<-uniroot(NegBinomGen, c(0, 0.99), R0=R0, k=k)$root
  R0star<-NegBinomR0star(R0,k)
  #need Q'"(z) evaluated at prob of extinction z < 1
  pgf2atextprob<-(R0^2)*((k+1)/k)*(1+(R0/k)*(1-extprob))^(-k-2)
  #Formula from Yan 2008:                                 
  varchain<-(extprob*pgf2atextprob+R0star*(1-R0star))/(1-R0star)^3
  return(varchain)
}

variance.chainNegBinomMix<-function(p, R0D,k, delta){ 
  #Argument: R0 > 1, R0A = (R0 - p R0^D) /(1-p)
  R0A<-R0D + delta
  extprob<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p, R0D=R0D, k=k, delta=delta)$root
  R0star<-NegBinomMixR0star(p,R0D,k, delta)
  #need Q'"(z) evaluated at prob of extinction z < 1
  pgf2atextprob<-p*(R0A^2)*((k+1)/k)*(1+(R0A/k)*(1-extprob))^(-k-2)+(1-p)*(R0D^2)*((k+1)/k)*(1+(R0D/k)*(1-extprob))^(-k-2)
  varchain<-(extprob*pgf2atextprob+R0star*(1-R0star))/(1-R0star)^3
  return(varchain)
  }



gNNB<-function(n, p, R0D, R0A, k){
  #inner function - negative binomial mixture
  #find its nth derivative
  q1 <- (1+R0A/k)
  q2 <- (1+R0D/k)
  
  if(n==1) {
    gn<-p*R0A*q1^(-k-1)+(1-p)*R0D*q2^(-k-1)
  }
  else if (n==2){
    gn<-p*((R0A^2)/k)*(k+1)*q1^(-k-2)+(1-p)*((R0D^2)/k)*(k+1)*q2^(-k-2)
  }
  else if (n>2) {
    pvec <- (k+1):(k+n-1)
  ppvec<-prod(pvec)
  #gn<-p*k*ppvec*((R0A/k)^n)*(1+R0A/k)^(-n-k)+(1-p)*(k)*ppvec*((R0D/k)^n)*(1+R0D/k)^(-n-k)
  gn<-p*((R0A/k)^n)*(k*ppvec)*q1^(-k-n)+(1-p)*((R0D/k)^n)*(k*ppvec)*q2^(-k-n) 
  }
  return(gn) 
}

fNrNB<-function(n, r, p, R0D, R0A, k){
  #outer function = pgf at s = 0
  #n is the power (a+b)^n, a = p*(1+R0A/k)^(-k), b=(1-p)*(1+R0D/k)^(-k)
  #r is the derivative required, r = 1, 2, ..., n-1
  #pgf of neg binom at s=0 is (1+R0/k)^(-k)
 fn<-(factorial(n)/factorial(n-r))*(p*(1+R0A/k)^(-k) +(1-p)*(1+R0D/k)^(-k))^(n-r)
}


nbinomchain<-function(j, R0, k){
  #cluster size distribution for negative binomial model
  #formula from Blumberg et al. 2013
  #Arguments: j=chain size, R0=mean, k=dispersion
  a<-R0/k
  x<-j-1
  y<-(a^x)/(1+a)^(k*j+x)
  g<-gamma(k*j+x)/(gamma(k*j)*gamma(j+1))
  return(g*y)
}

#prob cluster <=5
#better to find from the other matrix first?
#filter chains<=5
nbinommixturechain<-function(nend, p, R0D,k, delta){
  #generate cluster size distribution for a nbinom mixture, 
  #clusters of size 1, 2, ...., nend
  R0A<-R0D+delta
csize3end<-numeric(nend-2)

for(n in 3:nend){
  #for loop to calculate chain sizes greater than 2
nvec<-1:(n-1)
# gvec<-gNGeo(nvec, p, R0D, R0A) #need n-1 g derivatives

#compute the inner function 1, 2,..,n-1 derivatives of inner function 
gvec<-numeric(length(nvec))
for(i in 1:length(nvec)) gvec[i]<-gNNB(nvec[i],  p, R0D, R0A,k) #if statement so have to fill one by one
#The n-1th derivative of (a+b)^n evaluated at s = 0:

#use the derivatives to compute the  Bell polynomials vector of length n-1.

#Faa Di Bruno's formula = sum of outer function * Bell polys of the inner function derivatives
x1<-fNrNB(n, 1, p, R0D, R0A,k)*BellB(n-1,1,gvec) #r=1

for(i in 2:(n-1)) x1<-x1+fNrNB(n, i, p, R0D, R0A,k)*BellB(n-1,i,head(gvec,-(i-1)))

#vector of P(chain size = n)
csize3end[n-2]<-x1/factorial(n)
rm(x1)
}
#prob cluster size = 1 = Prob(cases=0)
csize1<-(p)*(1+R0A/k)^(-k)+(1-p)*(1+R0D/k)^(-k)
#prob cluster size = 2
csize2<-fNrNB(2, 1, p, R0D, R0A,k)*gNNB(1,p,R0D, R0A,k)/2

#chain size probabilities
csizemix<-c(csize1, csize2, csize3end)
return(csizemix)
}

ProbabilityChainSizeLessthanEqualC<-function(C, chainsizes){
  #chainsizes is a column vector of chain sizes
  #C =1, 2, ...,length(chainsizes)
  prob<-sum(chainsizes[1:C])
}

```


## Key questions

\begin{itemize}
\item Does the mechanistic addition of population structure induce qualitatively different outbreak patterns from a standard superspreading model? 
\item How does increasing the level of superspreading by a) changing the population structure e.g., by shifting the contact structure towards opportunistic contacts and away from regular contacts, and b) increasing the average number of successful contacts in the superspreading cohort affect heterogeneity in outbreak patterns, and what are the implications for containment?
\end{itemize}

## Model Assumptions

We assume that infected individuals can be divided into two disjoint groups - a fraction $p$ that contribute to transmission via superspreading, and the remaining fraction of the population $1-p$ that that do not contribute to superspreading transmission. In the superspreading cohort, the mean cumulative number of contacts leading to transmission of infection per infected individual per unit time is high at $\beta_1=\beta_D + \delta$, whereas in the non-superspreading group, it is low $\beta_2=\beta_D<\beta_1$. In both groups the contact process follows a Poisson distribution with mean $\beta_i$ $i,=1,2$. Then the contact process for the entire population is a finite Poisson mixture with probability distribution,
\begin{equation}
\text{number of cumulative contacts per infectious individual per unit time} \sim p 
\text{Poisson} (\beta_1 ) + (1-p) \text{Poisson} (\beta_2).
\end{equation}

In both groups, the infectious period is gamma distributed with mean $1/\gamma$ and coefficient of variation $1/\sqrt{k}$. The gamma distribution is a flexible distribution in that allows for right-skewed distributions (i.e., $k<1$) and distributions with a central tendency ($k>1$), with $k=1$ leading to the exponential distribution.  The probability generating function for the mixture follows
\begin{align}\label{eqn:nbinommix}
h(s) &= \int_0^\infty (p e^{\beta_1 x (s-1)}+(1-p)e^{\beta_2 x (s-1)}) \frac{(\gamma k)^k}{\Gamma(k)} x^{k-1}e^{-k \gamma x} dx \notag \\
&=p \frac{(\gamma k)^k}{(\gamma k+\beta_1(1-s))^k}+(1-p) \frac{(\gamma k)^k}{(\gamma k+\beta_2(1-s))^k} \notag \\
&=  p \left (\frac{1 }{(1+\frac{\beta_1}{\gamma k}(1-s))} \right )^k+(1-p) \left (\frac{1 }{(1+\frac{\beta_2}{\gamma k}(1-s))} \right )^k \notag \\
&=  \frac{p }{(1+\frac{R_0^A}{k}(1-s))^k}+  \frac{(1-p) }{(1+\frac{R_0^D}{k}(1-s))^k}
\end{align}

Therefore equation \eqref{eqn:nbinommix} shows that a finite mixture of negative binomial distributions models a combination of close contact transmission and superspreading. The probability mass function for the number of secondary infections per generation is 
\begin{equation}
\text{number of secondary infections} \sim p 
\text{Negative Binomial} (R_0^A,k ) + (1-p) \text{Negative Binomial} (R_0^D, k).
\end{equation}
The mean number of secondary infections is $R_0 = p R_0^A + (1-p)R_0^D$. 

## Probability mass functions for baseline and mixture models

Here we compare the probability mass functions of the mixture model ($R_0^D=1.1$, $p=0.1$, additional contacts $\delta =9$) with the base model for various values of $k$. The mean number of secondary infections for both models is $R_0 =2$. For the mixture models, the probability of no secondary infections is always greater than the negative binomial model with the same $R_0$ and $k$. As $k$ increases there is a greater central tendency in the number of secondary infections in the base model. 


```{r}
#parameters
k<-c(1/2, 1, 2, 4)
R0D<-1.1
R0<-2
p<-0.1
delta<-(R0-R0D)/p
R0A<-R0D+delta
```

```{r}
nend<-10 #max chain size
#base data is negative binomial model, R0=2, dispersion parameter = k
basedata<-matrix(NA, nrow=11, ncol=4) #cols are variables, rows are observations
for(i in 1:4) basedata[,i]<-dnbinom(0:10, size=k[i],mu=R0) #neg binom with R0=2
#mix data is negative binomial mixture model, R0=2, R0D, p as above dispersion parameter = k
mixdata<-matrix(NA, nrow=11, ncol=4)
for(i in 1:4) mixdata[,i]<-p*dnbinom(0:10, size=k[i],mu=R0D+delta)+(1-p)*dnbinom(0:10, size=k[i],mu=R0D)

#make a pmf data frame with columns = cases, proportions, dispersion parameters, model type
mdatapmf<-data.frame(Cases=0:10, Proportion=mixdata[,1], k=rep(k[1],11), Model=rep("mixture",11), R0=rep(2,nend+1), Distribution = rep("pmf",nend+1))
for(i in 2:4) mdatapmf<-bind_rows(mdatapmf,data.frame(Cases=0:10, Proportion=mixdata[,i], k=rep(k[i],11), Model=rep("mixture",11),R0=rep(2,nend+1), Distribution = rep("pmf",nend+1)))

bdatapmf<-data.frame(Cases=0:10, Proportion=basedata[,1], k=rep(k[1],11), Model=rep("base",11),R0=rep(2,nend+1), Distribution = rep("pmf",nend+1))
for(i in 2:4) bdatapmf<-bind_rows(bdatapmf,data.frame(Cases=0:10, Proportion=basedata[,i], k=rep(k[i],11), Model=rep("base",11),R0=rep(2,nend+1), Distribution = rep("pmf",nend+1)))

pmf<-bind_rows(bdatapmf, mdatapmf)

ggplot(pmf)+
   geom_point(aes(x = Cases, y = Proportion, colour=Model))+
  geom_line(aes(x = Cases, y = Proportion, colour=Model))+
  facet_wrap(~k, labeller=label_both)

```

## Corresponding chain size distributions 

Here we compare the chain size distributions of the mixture model ($R_0^D=1.1$, $p=0.1$, additional contacts $\delta =9$) with the base model for various values of $k$. The mean number of secondary infections for both models is $R_0 =2$. The chain size distribution is fatter tailed for the mixture models compared to the corresponding base models.


```{r}
nend<-10 #power ie (a+b)^n #nend = number of clusters, 1,...nend
csizemix<-matrix(NA, ncol=4, nrow=nend) # 4 cols for the 4 vals of k
for(j in 1:4){
  

kval<-k[j]
#generate the cluster distriubiton for the mixture distribution


csize3end<-numeric(nend-2)

for(n in 3:nend){
  #for loop to calculate chain sizes greater than 2
nvec<-1:(n-1)
# gvec<-gNGeo(nvec, p, R0D, R0A) #need n-1 g derivatives

#compute the inner function 1, 2,..,n-1 derivatives of inner function 
gvec<-numeric(length(nvec))
for(i in 1:length(nvec)) gvec[i]<-gNNB(nvec[i],  p, R0D, R0A,kval) #if statement so have to fill one by one
#The n-1th derivative of (a+b)^n evaluated at s = 0:

#use the derivatives to compute the  Bell polynomials vector of length n-1.

#Faa Di Bruno's formula = sum of outer function * Bell polys of the inner function derivatives
x1<-fNrNB(n, 1, p, R0D, R0A,kval)*BellB(n-1,1,gvec) #r=1

for(i in 2:(n-1)) x1<-x1+fNrNB(n, i, p, R0D, R0A,kval)*BellB(n-1,i,head(gvec,-(i-1)))

#vector of P(chain size = n)
csize3end[n-2]<-x1/factorial(n)
rm(x1)
}
#prob cluster size = 1 = Prob(cases=0)
csize1<-(p)*(1+R0A/kval)^(-kval)+(1-p)*(1+R0D/kval)^(-kval)
#prob cluster size = 2
csize2<-fNrNB(2, 1, p, R0D, R0A,kval)*gNNB(1,p,R0D, R0A,kval)/2

#chain size probabilities
csizemix[,j]<-c(csize1, csize2, csize3end)

}

```

```{r}
#Put into a data frame:
#data frame of chain size distribution for mixtures
clustermix<-data.frame(Chains=1:nend, Proportion=csizemix[,1], k=rep(k[1],nend), Model=rep("mixture",nend), R0=rep(2,nend), Distribution = rep("chain",nend)) 
for(i in 2:4) clustermix<-bind_rows(clustermix,data.frame(Chains=1:nend, Proportion=csizemix[,i], k=rep(k[i],nend), Model=rep("mixture",nend), R0=rep(2,nend), Distribution = rep("chain",nend)))

csizebase<-matrix(NA, nrow=nend, ncol=4)
for(i in 1:4) csizebase[,i]<-nbinomchain(1:10, R0, k[i])

clusterbase<-data.frame(Chains=1:nend, Proportion=csizebase[,1], k=rep(k[1],nend), Model=rep("base",nend), R0=rep(2,nend), Distribution = rep("chain",nend)) 
for(i in 2:4) clusterbase<-bind_rows(clusterbase,data.frame(Chains=1:nend, Proportion=csizebase[,i], k=rep(k[i],nend), Model=rep("base",nend), R0=rep(2,nend), Distribution = rep("chain",nend)))

clusterdistrib<-bind_rows(clusterbase, clustermix)
```


```{r}
#Plot shows that chain size distribution from a mixture is fatter tailed when $R_0>1$:
ggplot(clusterdistrib)+
   geom_point(aes(x = Chains, y = Proportion, colour=Model))+
  geom_line(aes(x = Chains, y = Proportion, colour=Model))+
  facet_wrap(~k, labeller=label_both)
```

## Statistics that show hallmarks of transmission heterogeneity

Hallmarks of heterogeneous transmission include:

\begin{itemize}
\item Greater variability in the number of secondary infections (fat tailed)
\item Smaller probability of major epidemics
\item Greater variability in chain sizes
\item Larger probability of observing no secondary infections and of observing small chains that go extinct
\end{itemize}

Here we study the coefficient of variation of the number of secondary infections, the probability of a major outbreak, the probability of observing a small transmission chain of less than or equal to 10 cases, and the mean and coefficient of variation of small chain sizes (conditioned on extinction). 

In each of the following, $p$ and $\delta$ are varied but $R_0 = R_0^D + \delta$ is fixed at $R_0 =2$. The following figures show that smaller values of $p$ (and larger values of $\delta$) lead to more heterogeneous epidemics, even if the dispersion parameter $k>1$. 

```{r}
#parameters
k<-c(1/2, 1, 2, 4)
R0D<-1.1
R0<-2
p<-seq(0.01, 1,0.01)
delta<-(R0-R0D)/p
R0A<-R0D+delta
```


```{r}
#Calculate mixture statistics:
nstat<-5
  mixstat<-array(NA, dim=c(length(k), length(p), nstat))
  for(j in 1:length(k)){
    kval<-k[j]
for(i in 1:length(p)){
  mixstat[j,i,1]<-varNegBinommixtureoffspring(p[i], R0D, kval, delta[i]) #variance of cases
  mixstat[j,i,2]<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p[i], R0D=R0D, k=kval, delta=delta[i])$root #extinction probability
  mixstat[j,i,3]<-NegBinomMixR0star(p[i], R0D, kval, delta[i])#R0* needed for mean chain size
  mixstat[j,i,4]<-variance.chainNegBinomMix(p[i], R0D, kval, delta[i]) #chain variance conditioned on extinction
  mixstat[j,i,5]<-ProbabilityChainSizeLessthanEqualC(C=10, nbinommixturechain(nend=10, p=p[i], R0D=R0D,k=kval, delta=delta[i])) #prob of observing cluster size <=10 (measure of stochastic burnout)
}  
  }

  #could include delta in the frame
#change 4 by pval by nstat array into  (4*pval) by nstat matrix/data frame
mstatdata<-data.frame(model="mixture", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p,delta=delta, k=rep(k[1], length(p)), CaseVariance=mixstat[1,,1], ExtinctionProbability=mixstat[1,,2],R0star=mixstat[1,,3],ChainVariance=mixstat[1,,4], ProbabilityChainSizeLessEqual10=mixstat[1,,5])
#stack each data frame corresponding to each k value below each other
for(i in 2:4) mstatdata<-bind_rows(mstatdata, data.frame(model="mixture", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p, delta=delta,k=rep(k[i], length(p)), CaseVariance=mixstat[i,,1], ExtinctionProbability=mixstat[i,,2],R0star=mixstat[i,,3],ChainVariance=mixstat[i,,4],ProbabilityChainSizeLessEqual10=mixstat[i,,5]))

#add CV cases, major outbreak probability, chain mean, CV chain
mstatdata<-mutate(mstatdata, CVCases=sqrt(CaseVariance)/R0, MajorOutbreak = 1-ExtinctionProbability, MeanChain = 1/(1-R0star),  CVChain=sqrt(ChainVariance)/MeanChain)
```



```{r}
#Calculate base statistics:
nstat<-5
  basestat<-matrix(NA, nrow=length(k), ncol=nstat)
  for(j in 1:length(k)){
    kval<-k[j]

  basestat[j,1]<-varNegBinomoffspring(R0, kval) #variance of cases
  basestat[j,2]<-uniroot(NegBinomGen, c(0, 0.99),   R0=R0, k=kval)$root #extinction probability
  basestat[j,3]<-NegBinomR0star(R0=R0, k=kval)#R0* needed for mean chain size
  basestat[j,4]<-variance.chainNegBinom(R0=R0, k=kval) #chain variance conditioned on extinction
  basestat[j,5]<-ProbabilityChainSizeLessthanEqualC(C=10, nbinomchain(1:10, R0, kval))
  }

# each stat repeated times so for every value of p, there is a corresponding base statistic
bstatdata<-data.frame(model="base", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p,delta=delta, k=rep(k[1], length(p)), CaseVariance=rep(basestat[1,1],length(p)), ExtinctionProbability=rep(basestat[1,2],length(p)),R0star=rep(basestat[1,3],length(p)),ChainVariance=rep(basestat[1,4],length(p)), ProbabilityChainSizeLessEqual10=rep(basestat[1,5],length(p)))
#put each data frame corresponding to each k value on top of each other
for(i in 2:4) bstatdata<-bind_rows(bstatdata, data.frame(model="base", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p, delta=delta,k=rep(k[i], length(p)), CaseVariance=rep(basestat[i,1],length(p)), ExtinctionProbability=rep(basestat[i,2],length(p)),R0star=rep(basestat[i,3],length(p)),ChainVariance=rep(basestat[i,4],length(p)), ProbabilityChainSizeLessEqual10=rep(basestat[i,5],length(p))))

#add CV cases, major outbreak probability, chain mean, CV chain
bstatdata<-mutate(bstatdata,CVCases=sqrt(CaseVariance)/R0, MajorOutbreak = 1-ExtinctionProbability, MeanChain = 1/(1-R0star),  CVChain=sqrt(ChainVariance)/MeanChain)
```

```{r}
statdata<-bind_rows(bstatdata, mstatdata)
```


## CV offspring distribution

There is greater variability in the number of secondary infections in the mixture model compared to the base model,even if $k>1$, with the highest variability for small dispersion parameter $k$, small $p$ and large number of additional contacts. 

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = CVCases, colour=model))+
  geom_line(aes(x = p, y = CVCases, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Probability of major outbreak

There is smaller probability of major epidemics in the mixture model compared to the base model, even if $k>1$, with the lowest probabilities for small dispersion parameter $k$, small $p$ and large number of additional contacts. 


```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = MajorOutbreak, colour=model))+
  geom_line(aes(x = p, y = MajorOutbreak, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Probability of observing a transmission chain of size <= 10

There is larger probability of observing small chains that go extinct in the mixture model compared to the base model, with the highest probabilities for small dispersion parameter $k$, small $p$ and large number of additional contacts. 

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = ProbabilityChainSizeLessEqual10, colour=model))+
  geom_line(aes(x = p, y = ProbabilityChainSizeLessEqual10, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Mean chain size

Mean chain sizes are larger in mixture models compared to the base model, even if $k>1$.

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = MeanChain, colour=model))+
  geom_line(aes(x = p, y = MeanChain, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## CV chain size

There is greater variability in chain sizes in the mixture model compared to the base model, even if $k>1$, with the highest coefficients of variation observed for small dispersion parameter $k$, small $p$ and large number of additional contacts. 

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = CVChain, colour=model))+
  geom_line(aes(x = p, y = CVChain, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Control activities

Here we will study the effect of two ways of reducing $R_0$:
\begin{itemize}
\item[(a)] decreasing the proportion $p$ of individuals in the population with high contact rate , which may
be considered to be the same as increasing the proportion of the population that self-isolate when sick or comply with
stay-at-home orders, with face covering mandates or other measures that reduce the chance of transmission;
\item [(b)] decreasing the number of additional contacts per individual in the superspreading cohort. 
\end{itemize}

In scenario (a), $p$ is reduced by multiplying it by by a factor $1-c$ (which in turn increasing the proportion in the non-superspreading cohort $1-p$ by a factor $1-c$) and in scenario (b), $\delta$ is reduced by multiplying it by by a factor $1-c$. In both control scenarios, effective $R_0$ is the same:
\begin{equation}
R_0^{\text{effective}} = R_0^D + (1-c) p \delta.
\end{equation}


```{r}
#parameters
control<-seq(0.01,1,0.01)
R0D<-1.1
R0<-2
pval<-0.1
deltaval<-(R0-R0D)/pval
R0A<-R0D+deltaval
kval<-1/2
```

```{r}
#Calculate mixture statistics:
nstat<-2
mixstatdel<-matrix(NA, nrow=length(control), ncol=nstat)

for(i in 1:length(control)){
  mixstatdel[i,1]<-varNegBinommixtureoffspring((1-control[i])*pval, R0D, kval, (1-control[i])*deltaval) #variance of cases
  mixstatdel[i,2]<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=pval, R0D=R0D, k=kval, delta=(1-control[i])*deltaval)$root #extinction probability
} 

mixstatp<-matrix(NA, nrow=length(control), ncol=nstat)

for(i in 1:length(control)){
  mixstatp[i,1]<-varNegBinommixtureoffspring((1-control[i])*pval, R0D, kval, deltaval) #variance of cases
  mixstatp[i,2]<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=(1-control[i])*pval, R0D=R0D, k=kval, delta=deltaval)$root #extinction probability
 
}  

cpstatdata<-data.frame(Model="Control on superspreading proportion", Control=control, CaseVariance=mixstatp[,1], ExtinctionProbability=mixstatp[,2])
cpstatdata<-mutate(cpstatdata, CVCases=sqrt(CaseVariance)/R0, MajorOutbreak = 1-ExtinctionProbability, p=(1-control)*pval, delta=rep(deltaval, length(control)))

cdeltastatdata<-data.frame(Model="Control on additional contacts",Control=control, CaseVariance=mixstatdel[,1], ExtinctionProbability=mixstatdel[,2])
cdeltastatdata<-mutate(cdeltastatdata, CVCases=sqrt(CaseVariance)/R0, MajorOutbreak = 1-ExtinctionProbability, p=rep(pval, length(control)), delta=(1-control)*deltaval)
                   
controlstat<-bind_rows(cpstatdata, cdeltastatdata)                   

```


Here we show that control actions that act to reduce the proportion of superspreaders $p$ in the population lead to greater heterogeneity in outbreaks (i.e., higher coefficient of variation in secondary infections and higher extinction probability) than reducing the number of additional contacts $\delta$. While reducing $p$ increases the probability of extinction, it also has the less desirable effect of increasing stochastic variation in outbreaks (e.g., coefficient of variation of secondary infections).

```{r}
ggplot(controlstat)+
   geom_point(aes(x = Control, y = CVCases, colour=Model))+
  geom_line(aes(x = Control, y = CVCases, colour=Model))
```
```{r}
ggplot(controlstat)+
   geom_point(aes(x = Control, y = ExtinctionProbability, colour=Model))+
  geom_line(aes(x = Control, y = ExtinctionProbability, colour=Model))
```

