---
title: "Negative binomial mixture branching process model of transmission"
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(BellB)
#library(actuar) #simulate mixture branching process
```

```{r global-options, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r}
R0fn<-function(p, R0D, R0A){
R0<- p*R0D+(1-p)*R0A
}

R0adel<-function(R0D, delta){
  R0A<-R0D + delta
}


NegBinomGen<-function(s, R0, k){
  #returns s = Q(s), where Q(s) = pgf of negative binomial
  gen<-1/(1 + (R0/k)*(1 - s))^k -s #solve prob generating fn = s for s. Note R0 > 1 otherwise prob of extinction = 1
}

NegBinomMixtureGen<-function(s, p, R0D,k, delta){
  #R0A = R0D + delta
  R0A<-R0D + delta
  gen<-p/(1 + (R0A/k)*(1 - s))^k+(1-p)/(1 + (R0D/k)*(1 - s))^k -s
}


mean.chain<-function(R0){ 
  #Argument R0 can be R0 if R0<1, if R0 >1, compute R0*
  meanchain<-1/(1-R0)
}

#variance of offspring function
varNegBinomoffspring<-function(R0,k){ 
  v<-R0*(1+R0/k)
}

#variance of offspring function
varNegBinommixtureoffspring<-function(p, R0D,k, delta){ 
  #variance is concave up quadratic function of p
  v<-R0D*(1+R0D/k) + p*delta*(1 + delta*(1-p) + (2*R0D+delta)/k)
}

NegBinomR0star<-function(R0=1.5, k){
  #R0 needs to be bigger than 1
  #R0* = Q'(z) evaluated at prob of extinction z < 1
  extprob<-uniroot(NegBinomGen, c(0, 0.99), R0=R0, k=k)$root
  R0star<-R0*(1+(R0/k)*(1-extprob))^(-k-1)
}

NegBinomMixR0star<-function(p, R0D,k, delta){
  R0A<-R0D + delta
  extprob<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p, R0D=R0D, k=k, delta=delta)$root
  R0star<-p*R0A*(1+(R0A/k)*(1-extprob))^(-k-1)+(1-p)*R0D*(1+(R0D/k)*(1-extprob))^(-k-1)
}


variance.chainNegBinom<-function(R0,k){ 
  #Argument: R0 > 1, dispersion parameter k
  #Find variance of chain conditioned on extinction
  #do i need a different function for R0<1??
  extprob<-uniroot(NegBinomGen, c(0, 0.99), R0=R0, k=k)$root
  R0star<-NegBinomR0star(R0,k)
  #need Q'"(z) evaluated at prob of extinction z < 1
  pgf2atextprob<-(R0^2)*((k+1)/k)*(1+(R0/k)*(1-extprob))^(-k-2)
  #Formula from Yan 2008:                                 
  varchain<-(extprob*pgf2atextprob+R0star*(1-R0star))/(1-R0star)^3
  return(varchain)
}

variance.chainNegBinomMix<-function(p, R0D,k, delta){ 
  #Argument: R0 > 1, R0A = (R0 - p R0^D) /(1-p)
  R0A<-R0D + delta
  extprob<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p, R0D=R0D, k=k, delta=delta)$root
  R0star<-NegBinomMixR0star(p,R0D,k, delta)
  #need Q'"(z) evaluated at prob of extinction z < 1
  pgf2atextprob<-p*(R0A^2)*((k+1)/k)*(1+(R0A/k)*(1-extprob))^(-k-2)+(1-p)*(R0D^2)*((k+1)/k)*(1+(R0D/k)*(1-extprob))^(-k-2)
  varchain<-(extprob*pgf2atextprob+R0star*(1-R0star))/(1-R0star)^3
  return(varchain)
  }



gNNB<-function(n, p, R0D, R0A, k){
  #inner function - negative binomial mixture
  #find its nth derivative
  q1 <- (1+R0A/k)
  q2 <- (1+R0D/k)
  
  if(n==1) {
    gn<-p*R0A*q1^(-k-1)+(1-p)*R0D*q2^(-k-1)
  }
  else if (n==2){
    gn<-p*((R0A^2)/k)*(k+1)*q1^(-k-2)+(1-p)*((R0D^2)/k)*(k+1)*q2^(-k-2)
  }
  else if (n>2) {
    pvec <- (k+1):(k+n-1)
  ppvec<-prod(pvec)
  #gn<-p*k*ppvec*((R0A/k)^n)*(1+R0A/k)^(-n-k)+(1-p)*(k)*ppvec*((R0D/k)^n)*(1+R0D/k)^(-n-k)
  gn<-p*((R0A/k)^n)*(k*ppvec)*q1^(-k-n)+(1-p)*((R0D/k)^n)*(k*ppvec)*q2^(-k-n) 
  }
  return(gn) 
}

fNrNB<-function(n, r, p, R0D, R0A, k){
  #outer function = pgf at s = 0
  #n is the power (a+b)^n, a = p*(1+R0A/k)^(-k), b=(1-p)*(1+R0D/k)^(-k)
  #r is the derivative required, r = 1, 2, ..., n-1
  #pgf of neg binom at s=0 is (1+R0/k)^(-k)
 fn<-(factorial(n)/factorial(n-r))*(p*(1+R0A/k)^(-k) +(1-p)*(1+R0D/k)^(-k))^(n-r)
}


nbinomchain<-function(j, R0, k){
  #cluster size distribution for negative binomial model
  #formula from Blumberg et al. 2013
  #Arguments: j=chain size, R0=mean, k=dispersion
  a<-R0/k
  x<-j-1
  y<-(a^x)/(1+a)^(k*j+x)
  g<-gamma(k*j+x)/(gamma(k*j)*gamma(j+1))
  return(g*y)
}

#prob cluster <=5
#better to find from the other matrix first?
#filter chains<=5
nbinommixturechain<-function(nend, p, R0D,k, delta){
  #generate cluster size distribution for a nbinom mixture, 
  #clusters of size 1, 2, ...., nend
  R0A<-R0D+delta
csize3end<-numeric(nend-2)

for(n in 3:nend){
  #for loop to calculate chain sizes greater than 2
nvec<-1:(n-1)
# gvec<-gNGeo(nvec, p, R0D, R0A) #need n-1 g derivatives

#compute the inner function 1, 2,..,n-1 derivatives of inner function 
gvec<-numeric(length(nvec))
for(i in 1:length(nvec)) gvec[i]<-gNNB(nvec[i],  p, R0D, R0A,k) #if statement so have to fill one by one
#The n-1th derivative of (a+b)^n evaluated at s = 0:

#use the derivatives to compute the  Bell polynomials vector of length n-1.

#Faa Di Bruno's formula = sum of outer function * Bell polys of the inner function derivatives
x1<-fNrNB(n, 1, p, R0D, R0A,k)*BellB(n-1,1,gvec) #r=1

for(i in 2:(n-1)) x1<-x1+fNrNB(n, i, p, R0D, R0A,k)*BellB(n-1,i,head(gvec,-(i-1)))

#vector of P(chain size = n)
csize3end[n-2]<-x1/factorial(n)
rm(x1)
}
#prob cluster size = 1 = Prob(cases=0)
csize1<-(p)*(1+R0A/k)^(-k)+(1-p)*(1+R0D/k)^(-k)
#prob cluster size = 2
csize2<-fNrNB(2, 1, p, R0D, R0A,k)*gNNB(1,p,R0D, R0A,k)/2

#chain size probabilities
csizemix<-c(csize1, csize2, csize3end)
return(csizemix)
}

ProbabilityChainSizeLessthanEqualC<-function(C, chainsizes){
  #chainsizes is a column vector of chain sizes
  #C =1, 2, ...,length(chainsizes)
  prob<-sum(chainsizes[1:C])
}

```


```{r}
#parameters
k<-c(1/2, 1, 2, 4)
R0D<-1.1
R0<-2
p<-0.1
delta<-(R0-R0D)/p
R0A<-R0D+delta
```

## Probability mass function



```{r}
nend<-10 #max chain size
#base data is negative binomial model, R0=2, dispersion parameter = k
basedata<-matrix(NA, nrow=11, ncol=4) #cols are variables, rows are observations
for(i in 1:4) basedata[,i]<-dnbinom(0:10, size=k[i],mu=R0) #neg binom with R0=2
#mix data is negative binomial mixture model, R0=2, R0D, p as above dispersion parameter = k
mixdata<-matrix(NA, nrow=11, ncol=4)
for(i in 1:4) mixdata[,i]<-p*dnbinom(0:10, size=k[i],mu=R0D+delta)+(1-p)*dnbinom(0:10, size=k[i],mu=R0D)

#make a pmf data frame with columns = cases, proportions, dispersion parameters, model type
mdatapmf<-data.frame(Cases=0:10, Proportion=mixdata[,1], k=rep(k[1],11), Model=rep("mixture",11), R0=rep(2,nend+1), Distribution = rep("pmf",nend+1))
for(i in 2:4) mdatapmf<-bind_rows(mdatapmf,data.frame(Cases=0:10, Proportion=mixdata[,i], k=rep(k[i],11), Model=rep("mixture",11),R0=rep(2,nend+1), Distribution = rep("pmf",nend+1)))

bdatapmf<-data.frame(Cases=0:10, Proportion=basedata[,1], k=rep(k[1],11), Model=rep("base",11),R0=rep(2,nend+1), Distribution = rep("pmf",nend+1))
for(i in 2:4) bdatapmf<-bind_rows(bdatapmf,data.frame(Cases=0:10, Proportion=basedata[,i], k=rep(k[i],11), Model=rep("base",11),R0=rep(2,nend+1), Distribution = rep("pmf",nend+1)))

pmf<-bind_rows(bdatapmf, mdatapmf)

ggplot(pmf)+
   geom_point(aes(x = Cases, y = Proportion, colour=Model))+
  geom_line(aes(x = Cases, y = Proportion, colour=Model))+
  facet_wrap(~k, labeller=label_both)

```

## Corresponding chain size distributions for the mixtures for each value of k:

```{r}
nend<-10 #power ie (a+b)^n #nend = number of clusters, 1,...nend
csizemix<-matrix(NA, ncol=4, nrow=nend) # 4 cols for the 4 vals of k
for(j in 1:4){
  

kval<-k[j]
#generate the cluster distriubiton for the mixture distribution


csize3end<-numeric(nend-2)

for(n in 3:nend){
  #for loop to calculate chain sizes greater than 2
nvec<-1:(n-1)
# gvec<-gNGeo(nvec, p, R0D, R0A) #need n-1 g derivatives

#compute the inner function 1, 2,..,n-1 derivatives of inner function 
gvec<-numeric(length(nvec))
for(i in 1:length(nvec)) gvec[i]<-gNNB(nvec[i],  p, R0D, R0A,kval) #if statement so have to fill one by one
#The n-1th derivative of (a+b)^n evaluated at s = 0:

#use the derivatives to compute the  Bell polynomials vector of length n-1.

#Faa Di Bruno's formula = sum of outer function * Bell polys of the inner function derivatives
x1<-fNrNB(n, 1, p, R0D, R0A,kval)*BellB(n-1,1,gvec) #r=1

for(i in 2:(n-1)) x1<-x1+fNrNB(n, i, p, R0D, R0A,kval)*BellB(n-1,i,head(gvec,-(i-1)))

#vector of P(chain size = n)
csize3end[n-2]<-x1/factorial(n)
rm(x1)
}
#prob cluster size = 1 = Prob(cases=0)
csize1<-(p)*(1+R0A/kval)^(-kval)+(1-p)*(1+R0D/kval)^(-kval)
#prob cluster size = 2
csize2<-fNrNB(2, 1, p, R0D, R0A,kval)*gNNB(1,p,R0D, R0A,kval)/2

#chain size probabilities
csizemix[,j]<-c(csize1, csize2, csize3end)

}

```

```{r}
#data frame of chain size distribution for mixtures
clustermix<-data.frame(Chains=1:nend, Proportion=csizemix[,1], k=rep(k[1],nend), Model=rep("mixture",nend), R0=rep(2,nend), Distribution = rep("chain",nend)) 
for(i in 2:4) clustermix<-bind_rows(clustermix,data.frame(Chains=1:nend, Proportion=csizemix[,i], k=rep(k[i],nend), Model=rep("mixture",nend), R0=rep(2,nend), Distribution = rep("chain",nend)))

csizebase<-matrix(NA, nrow=nend, ncol=4)
for(i in 1:4) csizebase[,i]<-nbinomchain(1:10, R0, k[i])

clusterbase<-data.frame(Chains=1:nend, Proportion=csizebase[,1], k=rep(k[1],nend), Model=rep("base",nend), R0=rep(2,nend), Distribution = rep("chain",nend)) 
for(i in 2:4) clusterbase<-bind_rows(clusterbase,data.frame(Chains=1:nend, Proportion=csizebase[,i], k=rep(k[i],nend), Model=rep("base",nend), R0=rep(2,nend), Distribution = rep("chain",nend)))

clusterdistrib<-bind_rows(clusterbase, clustermix)
```

Plot shows that chain size distribution from a mixture is fatter tailed when $R_0>1$:

```{r}
ggplot(clusterdistrib)+
   geom_point(aes(x = Chains, y = Proportion, colour=Model))+
  geom_line(aes(x = Chains, y = Proportion, colour=Model))+
  facet_wrap(~k, labeller=label_both)
```


```{r}
#parameters
k<-c(1/2, 1, 2, 4)
R0D<-1.1
R0<-2
p<-seq(0.01, 1,0.01)
delta<-(R0-R0D)/p
R0A<-R0D+delta
```


```{r}
#Calculate mixture statistics:
nstat<-5
  mixstat<-array(NA, dim=c(length(k), length(p), nstat))
  for(j in 1:length(k)){
    kval<-k[j]
for(i in 1:length(p)){
  mixstat[j,i,1]<-varNegBinommixtureoffspring(p[i], R0D, kval, delta[i]) #variance of cases
  mixstat[j,i,2]<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p[i], R0D=R0D, k=kval, delta=delta[i])$root #extinction probability
  mixstat[j,i,3]<-NegBinomMixR0star(p[i], R0D, kval, delta[i])#R0* needed for mean chain size
  mixstat[j,i,4]<-variance.chainNegBinomMix(p[i], R0D, kval, delta[i]) #chain variance conditioned on extinction
  mixstat[j,i,5]<-ProbabilityChainSizeLessthanEqualC(C=10, nbinommixturechain(nend=10, p=p[i], R0D=R0D,k=kval, delta=delta[i])) #prob of observing cluster size <=10 (measure of stochastic burnout)
}  
  }

  #could include delta in the frame
#change 4 by pval by nstat array into  (4*pval) by nstat matrix/data frame
mstatdata<-data.frame(model="mixture", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p,delta=delta, k=rep(k[1], length(p)), CaseVariance=mixstat[1,,1], ExtinctionProbability=mixstat[1,,2],R0star=mixstat[1,,3],ChainVariance=mixstat[1,,4], ProbabilityChainSizeLessEqual10=mixstat[1,,5])
#stack each data frame corresponding to each k value below each other
for(i in 2:4) mstatdata<-bind_rows(mstatdata, data.frame(model="mixture", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p, delta=delta,k=rep(k[i], length(p)), CaseVariance=mixstat[i,,1], ExtinctionProbability=mixstat[i,,2],R0star=mixstat[i,,3],ChainVariance=mixstat[i,,4],ProbabilityChainSizeLessEqual10=mixstat[i,,5]))

#add CV cases, major outbreak probability, chain mean, CV chain
mstatdata<-mutate(mstatdata, CVCases=sqrt(CaseVariance)/R0, MajorOutbreak = 1-ExtinctionProbability, MeanChain = 1/(1-R0star),  CVChain=sqrt(ChainVariance)/MeanChain)
```



```{r}
#Calculate base statistics:
nstat<-5
  basestat<-matrix(NA, nrow=length(k), ncol=nstat)
  for(j in 1:length(k)){
    kval<-k[j]

  basestat[j,1]<-varNegBinomoffspring(R0, kval) #variance of cases
  basestat[j,2]<-uniroot(NegBinomGen, c(0, 0.99),   R0=R0, k=kval)$root #extinction probability
  basestat[j,3]<-NegBinomR0star(R0=R0, k=kval)#R0* needed for mean chain size
  basestat[j,4]<-variance.chainNegBinom(R0=R0, k=kval) #chain variance conditioned on extinction
  basestat[j,5]<-ProbabilityChainSizeLessthanEqualC(C=10, nbinomchain(1:10, R0, kval))
  }

# each stat repeated times so for every value of p, there is a corresponding base statistic
bstatdata<-data.frame(model="base", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p,delta=delta, k=rep(k[1], length(p)), CaseVariance=rep(basestat[1,1],length(p)), ExtinctionProbability=rep(basestat[1,2],length(p)),R0star=rep(basestat[1,3],length(p)),ChainVariance=rep(basestat[1,4],length(p)), ProbabilityChainSizeLessEqual10=rep(basestat[1,5],length(p)))
#put each data frame corresponding to each k value on top of each other
for(i in 2:4) bstatdata<-bind_rows(bstatdata, data.frame(model="base", R0=rep(R0, length(p)), R0D=rep(R0D , length(p)),p=p, delta=delta,k=rep(k[i], length(p)), CaseVariance=rep(basestat[i,1],length(p)), ExtinctionProbability=rep(basestat[i,2],length(p)),R0star=rep(basestat[i,3],length(p)),ChainVariance=rep(basestat[i,4],length(p)), ProbabilityChainSizeLessEqual10=rep(basestat[i,5],length(p))))

#add CV cases, major outbreak probability, chain mean, CV chain
bstatdata<-mutate(bstatdata,CVCases=sqrt(CaseVariance)/R0, MajorOutbreak = 1-ExtinctionProbability, MeanChain = 1/(1-R0star),  CVChain=sqrt(ChainVariance)/MeanChain)
```

```{r}
statdata<-bind_rows(bstatdata, mstatdata)
```

In each of the following, $p$ and $\delta$ are varied but $R_0 = R_0^D + \delta$ is fixed at $R_0 =2$. The following figures show that smaller values of $p$ (and larger values of $\delta$) lead to more heterogeneous epidemics, even if $k>1$. Hallmarks of heterogeneous transmission include:

\begin{itemize}
\item Greater variability in the number of secondary infections (fat tailed)
\item Smaller probability of major epidemics
\item Greater variability in chain sizes
\item Larger probability of observing no secondary infections and small chains that go extinct
\end{itemize}

## CV offspring distribution

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = CVCases, colour=model))+
  geom_line(aes(x = p, y = CVCases, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Probability of major outbreak

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = MajorOutbreak, colour=model))+
  geom_line(aes(x = p, y = MajorOutbreak, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Probability of observing a transmission chain of size <= 10

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = ProbabilityChainSizeLessEqual10, colour=model))+
  geom_line(aes(x = p, y = ProbabilityChainSizeLessEqual10, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## Mean chain size

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = MeanChain, colour=model))+
  geom_line(aes(x = p, y = MeanChain, colour=model))+
  facet_wrap(~k, labeller=label_both)
```

## CV chain size

```{r}
ggplot(statdata)+
   geom_point(aes(x = p, y = CVChain, colour=model))+
  geom_line(aes(x = p, y = CVChain, colour=model))+
  facet_wrap(~k, labeller=label_both)
```
