---
title: "Superspreading negative binomial simulations"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r global-options, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}

R0fn<-function(p, R0D, R0A){
R0<- p*R0D+(1-p)*R0A
}

R0adel<-function(R0D, delta){
  R0A<-R0D + delta
}


NegBinomGen<-function(s, R0, k){
  #returns s = Q(s), where Q(s) = pgf of negative binomial
  gen<-1/(1 + (R0/k)*(1 - s))^k -s #solve prob generating fn = s for s. Note R0 > 1 otherwise prob of extinction = 1
}

NegBinomMixtureGen<-function(s, p, R0D,k, delta){
  #R0A = R0D + delta
  R0A<-R0D + delta
  gen<-p/(1 + (R0A/k)*(1 - s))^k+(1-p)/(1 + (R0D/k)*(1 - s))^k -s
}


mean.chain<-function(R0){ 
  #Argument R0 can be R0 if R0<1, if R0 >1, compute R0*
  meanchain<-1/(1-R0)
}

#variance of offspring function
varNegBinomoffspring<-function(R0,k){ 
  v<-R0*(1+R0/k)
}

#variance of offspring function
varNegBinommixtureoffspring<-function(p, R0D,k, delta){ 
  #variance is concave up quadratic function of p
  v<-R0D*(1+R0D/k) + p*delta*(1 + delta*(1-p) + (2*R0D+delta)/k)
}

NegBinomR0star<-function(R0=1.5, k){
  #R0 needs to be bigger than 1
  #R0* = Q'(z) evaluated at prob of extinction z < 1
  extprob<-uniroot(NegBinomGen, c(0, 0.99), R0=R0, k=k)$root
  R0star<-R0*(1+(R0/k)*(1-extprob))^(-k-1)
}

NegBinomMixR0star<-function(p, R0D,k, delta){
  R0A<-R0D + delta
  extprob<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p, R0D=R0D, k=k, delta=delta)$root
  R0star<-p*R0A*(1+(R0A/k)*(1-extprob))^(-k-1)+(1-p)*R0D*(1+(R0D/k)*(1-extprob))^(-k-1)
}


variance.chainNegBinom<-function(R0,k){ 
  #Argument: R0 > 1, dispersion parameter k
  #Find variance of chain conditioned on extinction
  #do i need a different function for R0<1??
  extprob<-uniroot(NegBinomGen, c(0, 0.99), R0=R0, k=k)$root
  R0star<-NegBinomR0star(R0,k)
  #need Q'"(z) evaluated at prob of extinction z < 1
  pgf2atextprob<-(R0^2)*((k+1)/k)*(1+(R0/k)*(1-extprob))^(-k-2)
  #Formula from Yan 2008:                                 
  varchain<-(extprob*pgf2atextprob+R0star*(1-R0star))/(1-R0star)^3
  return(varchain)
}

variance.chainNegBinomMix<-function(p, R0D,k, delta){ 
  #Argument: R0 > 1, R0A = (R0 - p R0^D) /(1-p)
  R0A<-R0D + delta
  extprob<-uniroot(NegBinomMixtureGen, c(0, 0.99),  p=p, R0D=R0D, k=k, delta=delta)$root
  R0star<-NegBinomMixR0star(p,R0D,k, delta)
  #need Q'"(z) evaluated at prob of extinction z < 1
  pgf2atextprob<-p*(R0A^2)*((k+1)/k)*(1+(R0A/k)*(1-extprob))^(-k-2)+(1-p)*(R0D^2)*((k+1)/k)*(1+(R0D/k)*(1-extprob))^(-k-2)
  varchain<-(extprob*pgf2atextprob+R0star*(1-R0star))/(1-R0star)^3
  return(varchain)
  }



gNNB<-function(n, p, R0D, R0A, k){
  #inner function - negative binomial mixture
  #find its nth derivative
  q1 <- (1+R0A/k)
  q2 <- (1+R0D/k)
  
  if(n==1) {
    gn<-p*R0A*q1^(-k-1)+(1-p)*R0D*q2^(-k-1)
  }
  else if (n==2){
    gn<-p*((R0A^2)/k)*(k+1)*q1^(-k-2)+(1-p)*((R0D^2)/k)*(k+1)*q2^(-k-2)
  }
  else if (n>2) {
    pvec <- (k+1):(k+n-1)
  ppvec<-prod(pvec)
  #gn<-p*k*ppvec*((R0A/k)^n)*(1+R0A/k)^(-n-k)+(1-p)*(k)*ppvec*((R0D/k)^n)*(1+R0D/k)^(-n-k)
  gn<-p*((R0A/k)^n)*(k*ppvec)*q1^(-k-n)+(1-p)*((R0D/k)^n)*(k*ppvec)*q2^(-k-n) 
  }
  return(gn) 
}

fNrNB<-function(n, r, p, R0D, R0A, k){
  #outer function = pgf at s = 0
  #n is the power (a+b)^n, a = p*(1+R0A/k)^(-k), b=(1-p)*(1+R0D/k)^(-k)
  #r is the derivative required, r = 1, 2, ..., n-1
  #pgf of neg binom at s=0 is (1+R0/k)^(-k)
 fn<-(factorial(n)/factorial(n-r))*(p*(1+R0A/k)^(-k) +(1-p)*(1+R0D/k)^(-k))^(n-r)
}


nbinomchain<-function(j, R0, k){
  #cluster size distribution for negative binomial model
  #formula from Blumberg et al. 2013
  #Arguments: j=chain size, R0=mean, k=dispersion
  a<-R0/k
  x<-j-1
  y<-(a^x)/(1+a)^(k*j+x)
  g<-gamma(k*j+x)/(gamma(k*j)*gamma(j+1))
  return(g*y)
}

#prob cluster <=5
#better to find from the other matrix first?
#filter chains<=5
nbinommixturechain<-function(nend, p, R0D,k, delta){
  #generate cluster size distribution for a nbinom mixture, 
  #clusters of size 1, 2, ...., nend
  R0A<-R0D+delta
csize3end<-numeric(nend-2)

for(n in 3:nend){
  #for loop to calculate chain sizes greater than 2
nvec<-1:(n-1)
# gvec<-gNGeo(nvec, p, R0D, R0A) #need n-1 g derivatives

#compute the inner function 1, 2,..,n-1 derivatives of inner function 
gvec<-numeric(length(nvec))
for(i in 1:length(nvec)) gvec[i]<-gNNB(nvec[i],  p, R0D, R0A,k) #if statement so have to fill one by one
#The n-1th derivative of (a+b)^n evaluated at s = 0:

#use the derivatives to compute the  Bell polynomials vector of length n-1.

#Faa Di Bruno's formula = sum of outer function * Bell polys of the inner function derivatives
x1<-fNrNB(n, 1, p, R0D, R0A,k)*BellB(n-1,1,gvec) #r=1

for(i in 2:(n-1)) x1<-x1+fNrNB(n, i, p, R0D, R0A,k)*BellB(n-1,i,head(gvec,-(i-1)))

#vector of P(chain size = n)
csize3end[n-2]<-x1/factorial(n)
rm(x1)
}
#prob cluster size = 1 = Prob(cases=0)
csize1<-(p)*(1+R0A/k)^(-k)+(1-p)*(1+R0D/k)^(-k)
#prob cluster size = 2
csize2<-fNrNB(2, 1, p, R0D, R0A,k)*gNNB(1,p,R0D, R0A,k)/2

#chain size probabilities
csizemix<-c(csize1, csize2, csize3end)
return(csizemix)
}

ProbabilityChainSizeLessthanEqualC<-function(C, chainsizes){
  #chainsizes is a column vector of chain sizes
  #C =1, 2, ...,length(chainsizes)
  prob<-sum(chainsizes[1:C])
}


NegBinmixturedeviate<- function(m, p, R0D, k, delta) {  
  #Returns a vector of m Poisson mixture deviates
  # Assumes a finite mixture of two Poisson distributions, with prob p and 1-p respectively, 0<p<1
  #1 = success = direct contact
  #0 = failure = aerosol
  #rbinom = one bernoulli trial for direct transmission
  #n = number of observations = 1
  #size = number of trials =1 (bernoulli) per observation
  #if success, deviate = rpois(R0D) otherwise aerosol
  R0A<-R0D +delta
  deviate<-numeric(m) #number of deviates
  for(i in 1:m){
deviate[i]<-ifelse(rbinom(1, size=1, prob = p)==1, rnbinom(1,mu=R0A, size=k), rnbinom(1,mu=R0D, size=k))
  }
  return(deviate)
}

NegBinmixturebranch<- function(n,p, R0D, k, delta) {
  #Arguments:
  #n = number of generations
  #Returns number of cases per generation as a vector
  #Uses Poissonmixturedeviate to generate random variates from mixture distribution
  
	z <- c(1,rep(0,n))#one infected case in generation 0 
	for (i in 2:(n+1)) {
z[i]<-ifelse(z[i-1]==0, 0, sum(NegBinmixturedeviate(z[i-1], p=p, R0D=R0D, k=k, delta=delta)))	  
			}
			return(z)
			}

NegBinbranch<- function(n,R0, k) {
  #Arguments:
  #n = number of generations
  #Returns number of cases per generation as a vector
  #Uses Poissonmixturedeviate to generate random variates from mixture distribution
  
	z <- c(1,rep(0,n))#one infected case in generation 0 
	for (i in 2:(n+1)) {
z[i]<-ifelse(z[i-1]==0, 0, sum(rnbinom(z[i-1],mu=R0,size=k)))	  
			}
			return(z)
			}

NegBinbranchw <- function(R0,k) {
  #returns n = the number of generations that it takes to have a cluster greater than 50 from a Poisson mixture offspring distribution
  
  n<-1
	z <- c(1,rep(0,n)) #n = number of generations after generation 0
	#one infected case in generation 0
	cluster<-sum(z)
	while(cluster<50){
	    i<-n+1
			z[i] <- sum(rnbinom(z[i-1],mu=R0,size=k))
			cluster<-cluster+z[i]
      n<-n+1
      if(z[i]==0){
        break #break out of loop of cluster dies out
      }
			}
			#return(c(z, cluster, n))
if(cluster<50)
	{n<-NA} #prob of extinction zinfinity predicts proportion of clusters that die out
	return(n)
			}

		
NegBinmixturebranchw<- function(p, R0D, k, delta) {  
#returns n = the number of generations that it takes to have a cluster greater than 50 from a Poisson mixture offspring distribution
R0A<-R0D+delta
  n<-1 #generation time
	z <- c(1,rep(0,n))
	#one infected case in generation 0 
	cluster<-sum(z)
	while(cluster<50){
	  #number of poisson rvs needed per case = z[i-1], then add together to get the number of new infections in that generation
	    i<-n+1
	    
	z[i]<-ifelse(z[i-1]==0, 0, sum(NegBinmixturedeviate(z[i-1], p=p, R0D=R0D, k=k, delta=delta)))	 

		cluster<-cluster+z[i]
      n<-n+1
      if(z[i]==0){
        break #if no cases break out of while loop
      }
			}
	#		return(c(z, cluster, n)) as a check
	if(cluster<50)
	{n<-NA}
	return(n)
			}


Poissonmixturebranchw2<- function(p, R0D, R0A) {  
#returns n = the number of generations that it takes to have a cluster greater than 50 from a Poisson mixture offspring distribution

  n<-1 #generation time
	z <- c(1,rep(0,n))
	#one infected case in generation 0 
	cluster<-sum(z)
	while(cluster<50){
	  #number of poisson rvs needed per case = z[i-1], then add together to get the number of new infections in that generation
	    i<-n+1
z[i] <- sum(rmixture(z[i-1], probs = c(p, 1-p), models = expression(rpois(R0D), rpois(R0A))))#params$R0D
		cluster<-cluster+z[i]
      n<-n+1
      if(z[i]==0){
        break #if no cases break out of while loop
      }
			}
	#		return(c(z, cluster, n)) as a check
	if(cluster<50)
	{n<-NA}
	return(n)
			}

```


```{r}
mydata<-data.frame(cases=NegBinmixturedeviate(100000, p=0.1, R0D=1.1, k=0.5,delta=(2-1.1)/0.1))
data<-data.frame(cases=rnbinom(100000, mu=2, size=0.5))
combdat<-bind_rows(list(Mixture=mydata,Baseline=data),
                           .id="model")
ggplot(combdat,aes(cases,y=stat(prop),fill=model))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position = 'identity') 
```

```{r}
count(mydata, cases)
```
If $p$ is small but $\delta$ is large, then the range of cases is large.

Now we wish to plot typical chains.

First lets look at the mean size of small (clusters < 10) and large clusters (>10).
The median clusters are typically bigger in the standard model than the mixture model but the mean is smaller, and mean chain sizes conditioned on extinction are similar in both models. If chopping the chain should condition on extinction/non-extinction?
```{r}
nsims<-1000
ngen=10
cluster<-numeric(nsims)
for(i in 1:nsims) cluster[i]<-sum(NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=0.5, delta=9)) 
c(mean(cluster[which(cluster>=10)]), range(cluster[which(cluster>=10)]))
median(cluster[which(cluster>=10)])
mean(cluster[which(cluster<10)])
```
```{r}
nsims<-1000
ngen=10
bcluster<-numeric(nsims)
for(i in 1:nsims) bcluster[i]<-sum(NegBinbranch(ngen, R0=2, k=0.5)) 
c(mean(bcluster[which(bcluster>=10)]), range(bcluster[which(bcluster>=10)]))
median(bcluster[which(bcluster>=10)])
mean(bcluster[which(bcluster<10)])
```
```{r}
nsims<-1000
ngen=20
bcluster<-numeric(nsims)
for(i in 1:nsims) bcluster[i]<-sum(NegBinbranch(ngen, R0=0.87, k=0.5)) 
mean(bcluster[which(bcluster>=10)])
median(bcluster[which(bcluster>=10)])
mean(bcluster[which(bcluster<10)])
```
```{r}
nsims<-1000
ngen=20
cluster<-numeric(nsims)
for(i in 1:nsims) cluster[i]<-sum(NegBinmixturebranch(ngen,p=0.1, R0D=0.5, k=0.5, delta=1)) 
mean(cluster[which(cluster>=10)])
median(cluster[which(cluster>=10)])
mean(cluster[which(cluster<10)])
```

```{r}
nsims<-10
ngen=10
sims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) sims[,i]<-NegBinmixturebranch(ngen,p=0.1, R0D=0.8, k=0.5, delta=1) 
sims
```
Probability of a major outbreak for mixture model is about 0.1.
```{r}
nsims<-10
ngen=10
sims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) sims[,i]<-NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=0.5, delta=9) 
sims
```
Probability of a major outbreak for standard model is about 0.4. Here four have taken off, but none are as large as the mixture that has taken off?
```{r}
nsims<-10
ngen=10
bsims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) bsims[,i]<-NegBinbranch(ngen, R0=2, k=0.5) 
bsims
```
Plot the mixture chains assuming $k=1/2:
```{r}
ngenmax<-4
simdf<-data.frame(Chain = sims[1:(ngenmax+1),1], ChainNum = "1", Generation=0:ngenmax)
chainnum<-c("2", "3", "4", "5", "6", "7", "8", "9", "10")
for(i in 2:nsims) simdf<-bind_rows(simdf, data.frame(Chain=sims[1:(ngenmax+1),i], ChainNum = chainnum[i-1], Generation=0:ngenmax))
```

```{r}
ggplot(simdf)+
   geom_point(aes(x = Generation, y = Chain, colour=ChainNum))+
  geom_line(aes(x = Generation, y = Chain, colour=ChainNum))
```
```{r}
nsims<-10
ngen=10
sims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) sims[,i]<-NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=2, delta=9) 
sims
```
Plot the mixture chains assuming $k=2:
```{r}
ngenmax<-8
simdf<-data.frame(Chain = sims[1:(ngenmax+1),1], ChainNum = "1", Generation=0:ngenmax)
chainnum<-c("2", "3", "4", "5", "6", "7", "8", "9", "10")
for(i in 2:nsims) simdf<-bind_rows(simdf, data.frame(Chain=sims[1:(ngenmax+1),i], ChainNum = chainnum[i-1], Generation=0:ngenmax))
```

```{r}
ggplot(simdf)+
   geom_point(aes(x = Generation, y = Chain,colour=ChainNum))+
  geom_line(aes(x = Generation, y = Chain, colour=ChainNum))+
  scale_colour_grey(start = 0, end = .9) +
  theme_bw()
```

```{r}
ggplot(simdf)+
   geom_point(aes(x = Generation, y = Chain, colour=ChainNum))+
  geom_line(aes(x = Generation, y = Chain, colour=ChainNum))
```
```{r}
nsims<-10
ngen=10
bsims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) bsims[,i]<-NegBinbranch(ngen, R0=2, k=2) 
bsims
```
```{r}
nsims<-10000
ngen=10
sims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) sims[,i]<-NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=2, delta=9) 
```

```{r}
ngenmax<-8
simdf<-data.frame(Chain = sims[1:(ngenmax+1),1], ChainNum = "1", Generation=0:ngenmax)
chainnum<-paste0("", 1:nsims)
for(i in 2:nsims) simdf<-bind_rows(simdf, data.frame(Chain=sims[1:(ngenmax+1),i], ChainNum = chainnum[i-1], Generation=0:ngenmax))
#View(simdf%>%filter(Generation==8 & Chain==0))
```



```{r}
#R0=4.5
nsims<-10000 #use while loops
kval<-0.5
R0<-2
R0D<-1.1
poissgen2<-numeric(nsims)
for(i in 1:length(poissgen2)) {
  poissgen2[i]<-NegBinbranchw(R0, kval)
}
#vary p =0.05, 0.1, 0.2
poissmixgen2<-numeric(nsims)
for(i in 1:length(poissmixgen2)) poissmixgen2[i]<-NegBinmixturebranchw(p=0.05, R0D, k=kval, delta=(R0-R0D)/0.05)

poissmixgen3<-numeric(nsims)
for(i in 1:length(poissmixgen3)) poissmixgen3[i]<-NegBinmixturebranchw(p=0.1, R0D, k=kval, delta=(R0-R0D)/0.1)
poissmixgen4<-numeric(nsims)
for(i in 1:length(poissmixgen4)) poissmixgen4[i]<-NegBinmixturebranchw(p=0.15, R0D, k=kval, delta=(R0-R0D)/0.15)
check<-c(length(na.omit(poissmixgen2)),length(na.omit(poissmixgen3)), length(na.omit(poissmixgen4)))
poissgen2<-na.omit(poissgen2)
poissmixgen2<-na.omit(poissmixgen2)
poissmixgen3<-na.omit(poissmixgen3)
poissmixgen4<-na.omit(poissmixgen4)

```

Median generation time same for all mixtures. Generation time has most variability for p=0.05, with 25% taking 11 generations or more to take off. 25% take 3 generations or less (most explosive for p =0.05, 0.1). 75% take 9 generations or less if p=0.1 (most explosive?)


```{r}


nsims2<-round(min(check),digits=-2) #round to the nearest 100
pval=rep(c("0.05", "0.1","0.15"), each=2*nsims2)
model=rep(c("baseline","mixture"),each=nsims2)
gentime<-c(poissgen2[1:nsims2], poissmixgen2[1:nsims2], poissgen2[1:nsims2],poissmixgen3[1:nsims2], poissgen2[1:nsims2], poissmixgen4[1:nsims2])
data2=data.frame(p=pval, model ,  GenerationTime=gentime)
 
# grouped boxplot
#p<-
ggplot(data2, aes(x=p, y=GenerationTime, fill=model)) + 
    geom_boxplot()

#ggsave("gentimeR0.pdf", p, height=4, width=5)
```

## Corresponding chain size distributions 

$k=2$, 4 generations (snapshot in time), plot only chains with size < 50:

```{r}
nsims<-10000
ngen=4
cluster<-numeric(nsims)
for(i in 1:nsims) cluster[i]<-sum(NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=2, delta=9)) 
bcluster<-numeric(nsims)
for(i in 1:nsims) bcluster[i]<-sum(NegBinbranch(ngen, R0=2, k=2)) 
mydata<-data.frame(ChainSize=cluster)
data<-data.frame(ChainSize=bcluster)
combdat<-bind_rows(list(Mixture=mydata,Baseline=data),
                           .id="model")
#plot only chains less than 50 (so prop is conditional)
ggplot(filter(combdat, ChainSize<50),aes(ChainSize,y=stat(prop),fill=model))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position = 'identity') 
```

Include only chains that have gone extinct after 10 generations:

```{r}
nsims<-1000
ngen=10
sims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) sims[,i]<-NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=2, delta=9)
mchains<-sims[,(sims[(ngen+1),]) == 0] #select only the columns with final generation cases = 0
bsims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) bsims[,i]<-NegBinbranch(ngen, R0=2, k=2) 
bchains<-bsims[,(bsims[(ngen+1),]) == 0]
mydata<-data.frame(ChainSize=colSums(mchains))
data<-data.frame(ChainSize=colSums(bchains))
combdat<-bind_rows(list(Mixture=mydata,Baseline=data),
                           .id="model")
ggplot(combdat,aes(ChainSize,y=stat(prop),fill=model))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position = 'identity') 
```

```{r}
dim(mchains)[2]/1000
dim(bchains)[2]/1000
summary(colSums(mchains))
summary(colSums(bchains))
```

The mixture chain size distribution is longer tailed than the baseline. The mixture chain size distribution contains many more chains that go extinct (minor outbreaks) than the baseline. The plot above conditions upon the probability of extinction, i.e. we see $P(Z=1)/z_\infty$, not $P(Z=1)$, e.g. for the standard model this is
```{r}
dnbinom(0, size=2, mu=2)/uniroot(NegBinomGen, c(0, 0.99), R0=2, k=2)$root
```

Plot that is not conditional upon extinction with largest chain size equal to the max of the chains that go extinct:
```{r}
nsims<-1000
ngen=10
sims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) sims[,i]<-NegBinmixturebranch(ngen,p=0.1, R0D=1.1, k=2, delta=9)
mchains<-sims[,(sims[(ngen+1),]) == 0] #select only the columns with final generation cases = 0
bsims<-matrix(NA, ncol=nsims, nrow=ngen+1)
for(i in 1:nsims) bsims[,i]<-NegBinbranch(ngen, R0=2, k=2) 
bchains<-bsims[,(bsims[(ngen+1),]) == 0]
mydata<-data.frame(ChainSize=colSums(sims))
data<-data.frame(ChainSize=colSums(bsims))
combdat<-bind_rows(list(Mixture=mydata,Baseline=data),
                           .id="model")
ggplot(combdat,aes(x=ChainSize,fill=model))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position = 'identity') +
  xlim(0, max(colSums(mchains)))

mixy<-count(mydata, ChainSize)
mixx<-mutate(mixy, ChainSizeProp=n/1000)
basey<-count(data, ChainSize)
basex<-mutate(basey, ChainSizeProp=n/1000)
combdat<-bind_rows(list(Mixture=mixx,Baseline=basex),
                           .id="model")
ggplot(combdat,aes(x=ChainSize, y=ChainSizeProp,colors=model))+
  geom_point()+
   scale_fill_manual(values=c("red","blue"))+
 # geom_bar(alpha=0.5,position = 'identity') +
  xlim(0, max(colSums(mchains)))
```


```{r}

pval<-c(0.05, 0.1, 0.15, 0.2)

#mix data is negative binomial mixture model, R0=2, R0D, p as above dispersion parameter = k
mixsim<-matrix(NA, nrow=10000, ncol=4)
for(i in 1:4) mixsim[,i]<-NegBinmixturedeviate(10000, p=pval[i], R0D=1.1, k=0.5,delta=(2-1.1)/pval[i])

#make a pmf data frame with columns = cases, proportions, dispersion parameters, model type
mdatasim<-data.frame(Cases=mixsim[,1], p=rep(pval[1],10000), Model=rep("mixture",10000), delta=(2-1.1)/pval[1])
for(i in 2:4) mdatasim<-bind_rows(mdatasim,data.frame(Cases=mixsim[,i], p=rep(pval[i],10000), Model=rep("mixture",10000), delta=(2-1.1)/pval[i]))

bdatasim<-data.frame(Cases=rnbinom(10000, mu=2, size=0.5),  p=rep(1,10000), Model=rep("base",10000), delta=(2-1.1))


sim05<-bind_rows(bdatasim, mdatasim)
sim05<-mutate(sim05, R0A=1.1+delta)
mdatasim<-mutate(mdatasim, R0A=1.1+delta)
```



```{r}
ggplot(mdatasim, aes(x = Cases, y=stat(prop), fill=p))+
  geom_bar(position = 'identity')+
  facet_wrap(~R0A, labeller=label_both)+
  xlim(-1,30)
```


```{r}
count(filter(mdatasim,R0A=="19.1"), Cases)
```


Compare negative binomial with $R_0=2$, $p=0.1$, $k=1/2$ with standard model, $R_0=2$, $k=1/2$:

```{r}
mydata<-data.frame(cases=NegBinmixturedeviate(100000, p=0.1, R0D=1.1, k=0.5,delta=(2-1.1)/0.1))
data<-data.frame(cases=rnbinom(100000, mu=2, size=0.5))
combdat<-bind_rows(list(Mixture=mydata,Baseline=data),
                           .id="model")
ggplot(combdat,aes(cases,y=stat(prop),fill=model))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position = 'identity') +
  ggtitle("k=1/2")
```

Compare negative binomial with $R_0=2$, $p=0.1$, $k=2$ with standard model, $R_0=2$, $k=2$. Similar to probability mass function for $k=2$ (below)

```{r}
mydata<-data.frame(cases=NegBinmixturedeviate(100000, p=0.1, R0D=1.1, k=2,delta=(2-1.1)/0.1))
data<-data.frame(cases=rnbinom(100000, mu=2, size=2))
combdat<-bind_rows(list(Mixture=mydata,Baseline=data),
                           .id="model")
ggplot(combdat,aes(cases,y=stat(prop),fill=model))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position = 'identity') +
  ggtitle("k=2")
```