
---
title: "Superspreading geometric mixture branching process"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
library(metR)
library(BellB)
library(actuar) #simulate mixture branching process
```

```{r global-options, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```




```{r}
R0fn<-function(p, R0D, R0A){
R0<- p*R0D+(1-p)*R0A
}

R0adel<-function(R0D, delta){
  R0A<-R0D + delta
}


GeometricGen<-function(s, R0=1.5){
  gen<-1/(1 + R0*(1 - s)) -s #solve prob generating fn = s for s=1/R0. Note R0 > 1 otherwise prob of extinction = 1
}

GeometricMixtureGen<-function(s, p, R0D,R0A){
  gen<-p/(1 + R0D*(1 - s))+(1-p)/(1 + R0A*(1 - s))-s
}

GeometricMixtureGenLimit<-function(s, p, R0D){
  gen<-p/(1 + R0D*(1 - s))-s
}

GeometricProportionNotInf<-function(R0=1.5){
  prop<-1/(1+R0)
}
GeometricMixProportionNotInf<-function(p, R0D,R0A){
  prop<-p/(1+R0D)+(1-p)/(1+R0A)
}
GeometricR0star<-function(R0=1.5){
  R0star<-R0/(1+(R0-1))^2
}
GeometricMixR0star<-function(p, R0D,R0A){
 extprob<-uniroot(GeometricMixtureGen, c(0, 0.99),  p=p, R0D=R0D, R0A=R0A)$root
  R0star<-p*R0D/(1+R0D*(1-extprob))^2+(1-p)*R0A/(1+R0A*(1-extprob))^2
  return(R0star)
}

mean.chain<-function(R0){ 
  meanchain<-1/(1-R0)
}

#variance.chainPoisson<-function(R0){ 
  #Argument: R0 > 1
 # extprob<-uniroot(PoissonGen, c(0, 0.9), R0=R0)$root
  #R0star<-PoissonR0star(R0)
  #varchain<-(extprob*(R0^2)*exp(R0*(extprob-1))+R0star*(1-R0star))/(1-R0star)^3
#}
#variance.chainPoissonMix<-function(p, R0D,R0A){ 
  #Argument: R0 > 1, R0A = (R0 - p R0^D) /(1-p)
 # extprob<-uniroot(PoissonMixtureGen, c(0, 0.99),  p=p, R0D=R0D, R0A=R0A)$root
  #R0star<-PoissonMixR0star(p, R0D,R0A)
  #varchain<-(extprob*((p*R0D^2)*exp(R0D*(extprob-1))+(1-p)*(R0A^2)*exp(R0A*(extprob-1)))+R0star*(1-R0star))/(1-R0star)^3
  #}

gNGeo<-function(n, p, R0D, R0A){
  gn<-factorial(n)*(p*(R0D^n)/(1+R0D)^(n+1)+(1-p)*(R0A^n)/(1+R0A)^(n+1))
}

fNrGeo<-function(n, r, p, R0D, R0A){
  #n is the power (a+b)^n, a = p*exp(-R0D), b=p*exp(-R0D)
  #r is the derivative required, r = 1, 2, ..., n-1
 fn<-(factorial(n)/factorial(n-r))*(p/(1+R0D) +(1-p)/(1+R0A))^(n-r)
}

#PoissonChain<-function(n, R0){
#  chain<-(exp(-n*R0)*(n*R0)^(n-1))/factorial(n)
#}



Geometricmixturedeviate<- function(m, p, R0D, R0A) {  
  #Returns a vector of m Poisson mixture deviates
  # Assumes a finite mixture of two Poisson distributions, with prob p and 1-p respectively, 0<p<1
  #1 = success = direct contact
  #0 = failure = aerosol
  #rbinom = one bernoulli trial for direct transmission
  #n = number of observations = 1
  #size = number of trials =1 (bernoulli) per observation
  #if success, deviate = rpois(R0D) otherwise aerosol
  deviate<-numeric(m) #number of deviates
  for(i in 1:m){
deviate[i]<-ifelse(rbinom(1, size=1, prob = p)==1, rgeom(1,1/(1+R0D)), rgeom(1,1/(1+R0A)))
  }
  return(deviate)
}

Geometricmixturebranch<- function(n,p, R0D, R0A) {
  #Arguments:
  #n = number of generations
  #Returns number of cases per generation as a vector
  #Uses Poissonmixturedeviate to generate random variates from mixture distribution
  
	z <- c(1,rep(0,n))#one infected case in generation 0 
	for (i in 2:(n+1)) {
z[i]<-ifelse(z[i-1]==0, 0, sum(Geometricmixturedeviate(z[i-1], p=p, R0D=R0D, R0A=R0A)))	  
			}
			return(z)
			}



Geometricbranch <- function(n,R0) {  ## Poisson
  #n = number of generations after generation 0
	z <- c(1,rep(0,n))
	#one infected case in generation 0
	for (i in 2:(n+1)) {
	  #number of poisson rvs needed per case = z[i-1], then add together to get the number of new infections in that generation
			z[i] <- sum(rpois(z[i-1],1/(1+R0)))
			}
			return(z)
			}
			

Geometicmixturebranch2<- function(n,p, R0D, R0A) { 
  ## requires rmixture from actuar package 
	z <- c(1,rep(0,n))
	for (i in 2:(n+1)) {
z[i] <- sum(rmixture(z[i-1], probs = c(p, 1-p), models = expression(rgeom(1/(1+R0D)), rgeom(1/(1+R0D)))))
			}
			return(z)
			}
			
clusterGeometric<-function(n, R0){
  #cluster size after n generations
  cluster<-sum(Geometricbranch(n,R0))
}

clusterGeometricmixture<-function(n, p, R0D, R0A){
  #cluster size after n generations
  cluster<-sum(Geometricmixturebranch(n,p, R0D, R0A))
}

clusterGeometricmixture2<-function(n, p, R0D, R0A){
  #cluster size after n generations
  cluster<-sum(Geometricmixturebranch2(n,p, R0D, R0A))
}

Geometricbranchw <- function(R0) {
  #returns n = the number of generations that it takes to have a cluster greater than 50 from a Poisson mixture offspring distribution
  
  n<-1
	z <- c(1,rep(0,n)) #n = number of generations after generation 0
	#one infected case in generation 0
	cluster<-sum(z)
	while(cluster<50){
	    i<-n+1
			z[i] <- sum(rgeom(z[i-1],1/(1+R0)))
			cluster<-cluster+z[i]
      n<-n+1
      if(z[i]==0){
        break #break out of loop of cluster dies out
      }
			}
			#return(c(z, cluster, n))
if(cluster<50)
	{n<-NA} #prob of extinction zinfinity predicts proportion of clusters that die out
	return(n)
			}

		
Geometricmixturebranchw<- function(p, R0D, R0A) {  
#returns n = the number of generations that it takes to have a cluster greater than 50 from a Poisson mixture offspring distribution

  n<-1 #generation time
	z <- c(1,rep(0,n))
	#one infected case in generation 0 
	cluster<-sum(z)
	while(cluster<50){
	  #number of poisson rvs needed per case = z[i-1], then add together to get the number of new infections in that generation
	    i<-n+1
	    
	z[i]<-ifelse(z[i-1]==0, 0, sum(Geometricmixturedeviate(z[i-1], p=p, R0D=R0D, R0A=R0A)))	 

		cluster<-cluster+z[i]
      n<-n+1
      if(z[i]==0){
        break #if no cases break out of while loop
      }
			}
	#		return(c(z, cluster, n)) as a check
	if(cluster<50)
	{n<-NA}
	return(n)
			}


Geometricmixturebranchw2<- function(p, R0D, R0A) {  
#returns n = the number of generations that it takes to have a cluster greater than 50 from a Poisson mixture offspring distribution

  n<-1 #generation time
	z <- c(1,rep(0,n))
	#one infected case in generation 0 
	cluster<-sum(z)
	while(cluster<50){
	  #number of poisson rvs needed per case = z[i-1], then add together to get the number of new infections in that generation
	    i<-n+1
z[i] <- sum(rmixture(z[i-1], probs = c(p, 1-p), models = expression(rgeom(1/(1+R0D)), rgeom(1/(1+R0A)))))#params$R0D
		cluster<-cluster+z[i]
      n<-n+1
      if(z[i]==0){
        break #if no cases break out of while loop
      }
			}
	#		return(c(z, cluster, n)) as a check
	if(cluster<50)
	{n<-NA}
	return(n)
			}

```

## Offspring Distribution


```{r}
data<-data.frame(cases=rmixture(1000, probs = c(4, 1), models = expression(rgeom(prob=1/(1+1.5)), rgeom(prob=1/(1+16.5)))))

data2<-data.frame(cases=rgeom(1000,prob=1/(1+4.5)))
combdat<-bind_rows(list(Mixture=data,Baseline=data2),
                           .id="model")
ggplot(combdat,aes(cases,fill=model,y = stat(prop)))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5, position ='identity') 

```
```{r}
ggplot(filter(combdat, cases<50),aes(cases,fill=model,y = stat(prop)))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5, position ='identity') 
```


```{r}
#rewrite without using actuar package
#R0 = 4.5
data<-data.frame(cases=rmixture(10000, probs = c(0.8, 0.2), models = expression(rgeom(prob=1/(1+1.5)), rgeom(prob=1/(1+16.5)))))
data2<-data.frame(cases=rgeom(10000, prob=1/(1+4.5)))
combdat<-bind_rows(list(GeometricMixture=data,GeometricBaseline=data2),
                           .id="model")
combdat2<-mutate(combdat, R0=4.5)
#R0 = 2
data1<-data.frame(cases=rmixture(10000, probs = c(0.8, 0.2), models = expression(rgeom(prob=1/(1+1.5)), rgeom(prob=1/(1+4)))))
data3<-data.frame(cases=rgeom(10000, prob=1/(1+2)))
combdat3<-bind_rows(list(GeometricMixture=data1,GeometricBaseline=data3),
                           .id="model")
combdat4<-mutate(combdat3, R0=2)
#combine together
combdatf<-bind_rows(combdat2, combdat4)

plot<-ggplot(combdatf,aes(cases,fill=model, y = stat(prop)))+
   scale_fill_manual(values=c("red","blue"))+
  geom_bar(alpha=0.5,position='identity')+
  facet_wrap(~R0, labeller=label_both)
plot
ggsave("GeoCaseDistribution.pdf", plot, height=3, width=5)
```

\textbf{Figure.} The case distribution generated by a geometric mixture compared with baseline geometric distributions with the same $R_0$. The geometric mixture offspring distribution with $p=0.8$, $R_0^D=1.5$ and $R_0^A=4$ (mean $R_0=2$) and the geometric mixture offspring distribution with $p=0.8$, $R_0^D=1.5$ and $R_0^A=16.5$ (mean $R_0=4.5$) are both right skewed. The case distribution generated by the respective baseline geometric offspring distributions are right skewed but have shorter tails. To generate the barplots, we used 10,000 realizations of each offspring distribution. Purple indicates overlap of the baseline and mixture distributions.

## Probability at least one becomes infected

```{r}
R0D=1.5;
R0start=1.6;
R0end=10;
pseq<-seq(0.6, 0.9, 0.1)
delta6<-seq((R0start-R0D)/(1-pseq[1]),(R0end-R0D)/(1-pseq[1]), length.out=100)
delta7<-seq((R0start-R0D)/(1-pseq[2]),(R0end-R0D)/(1-pseq[2]),length.out=100)
delta8<-seq((R0start-R0D)/(1-pseq[3]),(R0end-R0D)/(1-pseq[3]), length.out=100)
delta9<-seq((R0start-R0D)/(1-pseq[4]),(R0end-R0D)/(1-pseq[4]), length.out=100)
R0vals6<-R0D+delta6*(1-pseq[1])
R0vals7<-R0D+delta7*(1-pseq[2])
R0vals8<-R0D+delta8*(1-pseq[3])
R0vals9<-R0D+delta9*(1-pseq[4])
prop.notinf<-GeometricProportionNotInf(R0vals6)
prop.notinfp6<-GeometricMixProportionNotInf(p=0.6, R0D=1.5, R0A=R0adel( R0D=1.5, delta=delta6))
prop.notinfp7<-GeometricMixProportionNotInf(p=0.7, R0D=1.5, R0A=R0adel( R0D=1.5, delta=delta7))
prop.notinfp8<-GeometricMixProportionNotInf(p=0.8, R0D=1.5, R0A=R0adel( R0D=1.5, delta=delta8))
prop.notinfp9<-GeometricMixProportionNotInf(p=0.9, R0D=1.5, R0A=R0adel( R0D=1.5, delta=delta9))
plot(prop.notinf~R0vals6, xlab="R0", ylab="Proportion not infected (P(X=0))", ylim=c(0,0.4))
points(prop.notinfp6~R0vals6, col="blue")
points(prop.notinfp7~R0vals7, col="red")
points(prop.notinfp8~R0vals8, col="cyan")
points(prop.notinfp9~R0vals9, col="purple")
legend(2,0.2,c("Baseline", "Aerosol 40%", "Aerosol 30%","Aerosol 20%","Aerosol 10%"), col=c('black', 'blue', 'red', 'cyan', 'purple'), lty=1, bty='n', pch=1)
```
\textbf{Figure.} For each value of $R_0$, the mixture models predict that a higher proportion of individuals remain uninfected than the baseline Poisson model, and the proportion decreases with increasing aerosol transmission. Geometric mixture models saturate more slowly than corresponding Poisson models.


## Probability of major epidemic


```{r}

R0Aval<-0.5+seq(0.1, 145, 0.1)
R0<-0.5+0.2*seq(0.1, 145, 0.1)
index<-which(R0>1)[1]
prob.extlimit<-uniroot(GeometricMixtureGenLimit, c(0, 0.999),  p=0.8, R0D=0.5)$root
prob.ext<-numeric(length(R0Aval))
prob.ext[1:(index)]<-1 #goes extinct if R0 <=1
for(i in (index):length(R0Aval)){
  prob.ext[i]<-uniroot(GeometricMixtureGen, c(0, 0.999999999),  p=0.8, R0D=0.5, R0A=R0Aval[i])$root
}

test_data <-
  data.frame(
    limit = 1-prob.extlimit,
     mixture= 1-prob.ext,
  R0A = R0Aval)
  

test_data_long <- melt(test_data, id="R0A")  # convert to long format

ggplot(data=test_data_long,
       aes(x=R0A, y=value, colour=variable)) +
    geom_line() +
  labs(x="Aerosol R0", y="Probability of major epidemic")+
    scale_color_discrete(name = "Model", labels = c("Limit", "Mixture"))
```
\textbf{Figure.} Approaches the limit of the solution of  $z = p/(1+R_0^D(1-z)$ much more slowly relative to Poisson epidemic (the equation has two solutions; one lies between 0 and 1).

```{r}
R0D=1.5
prob.ext<-numeric(length(R0vals6))
prob.extp6<-numeric(length(R0vals6))
prob.extp7<-numeric(length(R0vals7))
prob.extp8<-numeric(length(R0vals8))
prob.extp9<-numeric(length(R0vals9))
for(i in 1:length(R0vals6)){
  prob.ext[i]<-uniroot(GeometricGen, c(0, 0.99), R0=R0vals6[i])$root
  prob.extp6[i]<-uniroot(GeometricMixtureGen, c(0, 0.999),  p=0.6, R0D=1.5, R0A=R0adel(R0D=1.5, delta=delta6[i]))$root
  prob.extp7[i]<-uniroot(GeometricMixtureGen, c(0, 0.99),  p=0.7, R0D=1.5, R0A=R0adel(R0D=1.5, delta=delta7[i]))$root
  prob.extp8[i]<-uniroot(GeometricMixtureGen, c(0, 0.99),  p=0.8, R0D=1.5, R0A=R0adel(R0D=1.5, delta=delta8[i]))$root
  prob.extp9[i]<-uniroot(GeometricMixtureGen, c(0, 0.99),  p=0.9, R0D=1.5, R0A=R0adel(R0D=1.5, delta=delta9[i]))$root
}
plot(prob.ext~R0vals6, xlab="R0", ylab="Probability of extinction", ylim=c(0,0.65), main ="Probability of extinction of a minor outbreak")
points(prob.extp6~R0vals6, col="blue")
points(prob.extp7~R0vals7, col="red")
points(prob.extp8~R0vals8, col="cyan")
points(prob.extp9~R0vals9, col="purple")
legend(1.5,0.25,c("Baseline", "Aerosol 40%", "Aerosol 30%","Aerosol 20%","Aerosol 10%"), col=c('black', 'blue', 'red', 'cyan', 'purple'), lty=1, bty='n', pch=1, cex=0.6)
```

## First generation to 50 cases


```{r}
nsims<-1000 #use while loops
geomgen2<-numeric(nsims)
for(i in 1:length(geomgen2)) {
  geomgen2[i]<-Geometricbranchw(2)
}
geommixgen2<-numeric(nsims)
for(i in 1:length(geommixgen2)) geommixgen2[i]<-Geometricmixturebranchw(0.8, 1.5, 4)
geomgen3<-numeric(nsims)
for(i in 1:length(geomgen3)) geomgen3[i]<-Geometricbranchw(3)
geommixgen3<-numeric(nsims)
for(i in 1:length(geommixgen3)) geommixgen3[i]<-Geometricmixturebranchw(0.8, 1.5, 9)
geomgen4<-numeric(nsims)
for(i in 1:length(geomgen4)) geomgen4[i]<-Geometricbranchw(4)
geommixgen4<-numeric(nsims)
for(i in 1:length(geommixgen4)) geommixgen4[i]<-Geometricmixturebranchw(0.8, 1.5, 14)
check<-c(length(na.omit(geomgen2)),length(na.omit(geommixgen2)),length(na.omit(geomgen3)),length(na.omit(geommixgen3)),length(na.omit(geomgen4)), length(na.omit(geommixgen4)))
geomgen2<-na.omit(geomgen2)
geommixgen2<-na.omit(geommixgen2)
geomgen3<-na.omit(geomgen3)
geommixgen3<-na.omit(geommixgen3)
geomgen4<-na.omit(geomgen4)
geommixgen4<-na.omit(geommixgen4)

```

```{r}


nsims2<-round(min(check),digits=-2) #round to the nearest 100
R0=rep(c("2", "3","4"), each=2*nsims2)
model=rep(c("geometric baseline ","geometric mixture"),each=nsims2)
gentime<-c(geomgen2[1:nsims2], geommixgen2[1:nsims2],geomgen3[1:nsims2], geommixgen3[1:nsims2], geomgen4[1:nsims2], geommixgen4[1:nsims2])
gentimedata=data.frame(R0, model ,  gentime)
 
# grouped boxplot
p<-ggplot(gentimedata, aes(x=R0, y=gentime, fill=model)) + 
    geom_boxplot()
p
ggsave("gentimegeoR0.pdf", p, height=4, width=5)
```
\textbf{Figure.} More variability in geometric epidemics than Poisson epidemics. When $R_0 = 4$ about 21\% of epidemics have generation times between 2 and 3 (i.e. are explosive) and about 8\% have generation times greater or equal than 7 (i.e are relatively slow). 


```{r}
table(geommixgen4[1:nsims2])
summary(geommixgen4[1:nsims2])
table(geommixgen4[1:nsims2])/nsims2
sum((table(geommixgen4[1:nsims2])/nsims2)[5:9])
```


## Cluster size distribution conditioned on extinction

Need to have cluster sizes greater than 10 to capture the theoretical prediction $1/(1-R_0^*)$ for the mean chain size:
```{r}
R0star<-GeometricMixR0star(p=0.8, R0D=1.5, R0A=4)
theor.mean.chain<-mean.chain(R0star)
theor.mean.chain
#variance

```
Maybe should reduce the clusters to < 16

```{r}

#simulated cluster sizes for the Poisson mixture distribution:
clusterM<-numeric(100000)
for(i in 1:length(clusterM) )clusterM[i]<-clusterGeometricmixture(6, p=0.8, R0D=1.5, R0A=4) #how many go extinct after n=6 generations?
clusterM<-data.frame(clusters=clusterM)
clusterM2<-clusterM%>%filter(clusters<21)
dim(clusterM2) #use to obtain the proportion that have gone extinct

extprob1<-uniroot(GeometricGen, c(0, 0.99), R0=2)$root
extprobM<-uniroot(GeometricMixtureGen, c(0, 0.99),  p=0.8, R0D=1.5, R0A=4)$root

#generate the cluster distriubiton for the mixture distribution
p2<-0.8
R0D2<-1.5
R0A2<-4

nend<-20#power ie (a+b)^n
csize32<-numeric(nend-2)
for(n in 3:nend){
nvec<-1:(n-1)
gvec<-gNGeo(nvec, p2, R0D2, R0A2) #need n-1 g derivatives

#The n-1th derivative of (a+b)^n evaluated at s = 0:
#the for loop should work for chain sizes greater than 2
x1<-fNrGeo(n, 1, p2, R0D2, R0A2)*BellB(n-1,1,gvec)
for(k in 2:(n-1)) x1<-x1+fNrGeo(n, k, p2, R0D2, R0A2)*BellB(n-1,k,head(gvec,-(k-1)))

#vector of P(chain size = n)
csize32[n-2]<-x1/factorial(n)
rm(x1)
}
csize12<-p2/(1+R0D2)+(1-p2)/(1+R0A2)
csize22<-fNrGeo(2, 1, p2, R0D2, R0A2)*gNGeo(1,p2,R0D2, R0A2)/2
#compute chain size distribution
csize2<-c(csize12, csize22, csize32)
#cluster sizes conditional on extinction
clusterthM<-data.frame(n=1:20, p=csize2/extprobM) 


plot<-ggplot() + 
  geom_bar(data = clusterM2, aes(x = clusters, y = stat(prop))) + 
  geom_point(data = clusterthM, aes(x=n, y=p))+
  geom_line(data = clusterthM, aes(x=n, y=p))
plot
ggsave("geoclustersize.pdf", plot, width=5, height=4)

```

\textbf{Figure.} Geometric mixture cluster size distribution obtained from simulation (gray bars) compared to the theoretical prediction (points). We generated 100000 simulations of a geometric mixture branching process with $p=0.8$, $R_0^D = 1.5$ and $R_0^A=4$, retaining those that went extinct within 6 generations ($n=56150$). Points are the theoretical predictions for the probability of observing a cluster of size $y=1,2,\dots, 10$ generated using equations \eqref{eqn:cluster} and \eqref{eqn:bell} conditioned on the probability of extinction, $z_\infty = 0.548$.

The theoretical cluster size distribution is given by:
```{r}
mutate(clusterthM, n*p)
```

Use the theoretical distribution to calculate the mean chain size, the sum of the n*p column (mean = $\sum_i y_i P(Y=y_i$)). Note that the sum of the probabilities is close to 1, but less than one, so the calculation yields an underestimate of the mean:
```{r}
colSums(mutate(clusterthM, n*p))
```

